Determining sequence-dependent DNA oligonucleotide hybridization and dehybridization mechanisms using coarse-grained molecular simulation, Markov state models, and infrared spectroscopy

### 2021-05-17 manuscript.tex
acs style, citation abbreviations, figure 3 label

%add refs to SI
%\usepackage{xr}
%\externaldocument{SI}

%\usepackage[hidelinks]{hyperref}
%\usepackage{algorithm}
%\usepackage{algpseudocode}
%\algdef{SE}[DOWHILE]{Do}{doWhile}{\algorithmicdo}[1]{\algorithmicwhile\ #1}

@@

\newcommand*\mycommand[1]{\texttt{\emph{#1}}}

\newcommand{\state}[1]{$\mathcal{S}_{#1}$}
\newcommand{\Expect}[1]{\mathrm{E}\left[#1\right]}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}

\newcommand*{\rood}[1]{{\color{red}{#1}}}
\newcommand*{\bood}[1]{{\color{blue}{#1}}}
\newcommand*{\blauw}[1]{\textcolor{blue}{#1}}
\newcommand*{\groen}[1]{\textcolor{green}{#1}}

\newcommand*{\addref}[1]{\textcolor{red}{\{ADD REF: #1\}}}
\newcommand*{\noter}[1]{\textcolor{red}{[[#1]]}}		% notes 
\newcommand*{\noteb}[1]{\textcolor{blue}{[[#1]]}}		% notes 
%\newcommand*{\noter}[1]{}					% notes off

@@

Markov state models (MSMs) are a powerful approach to infer long-time kinetic models from short molecular simulation trajectories \citep{Sengupta2019AutomatedSelf-assembly, Pande2010EverythingAsk, Chodera2014MarkovDynamics, Husic2018MarkovScience, Wehmeyer2019IntroductionSoftware} that we employ to construct high-resolution sequence-dependent kinetic models of DNA hybridization and dissociation. In brief, a MSM extracts from simulation trajectories an ensemble of long-lived metastable macrostates, their equilibrium occupancy populations, and the equilibrium probability fluxes between them. In this manner, they provide an interpretable and predictive model of the system thermodynamics and long-time kinetics. MSMs have recently been implemented to study the hybridization mechanism of one particular 14-mer DNA oligomer, but determining the sequence-dependent kinetics and mechanisms was not the focus of this study.\citep{Jin2019} An energy disconnectivity graph-based approach was used to interrogate the differences in hybridization rates and mechanisms between GGGGGG and GCGCGC hexamers to reveal strong deviations from ``all-or-nothing'' behaviors and the importance of zippering and slithering mechanisms.~\citep{Xiao2019} Kinetic models were constructed not from the dynamical trajectories of the molecular model, but by estimating rate constants between local minima were estimated using a transition state theory approximation. A recent application of MSMs to the long-time dynamics of short RNA oligonucleotides revealed stacking time scales to be highly sequence dependent.~\citep{Pinamonti2017} In this work, MSMs were constructed for each of the four DNA sequences from the 40$\times$25 $\mu$s simulation trajectories following the six-step protocol detailed in Ref.~\citep{Sidky}: (i) trajectory featurization, (ii) dimensionality reduction, (iii) microstate clustering and microstate transition matrix inference, (iv) macrostate clustering and macrostate transition matrix inference. Calculations were performed using the PyEMMA software package. \citep{Scherer2015PyEMMAModels}

@@

\textbf{Dimensionality reduction.} The featurized trajectories were then projected into a low-dimensional space in preparation for microstate clustering. The standard approach to doing so is to employ time-lagged independent components analysis (tICA) to learn a linear projection into a low-dimensional embedding that maximally preserves the kinetic variance in the data.~\citep{Perez-Hernandez2013IdentificationConstruction, Schwantes2013ImprovementsNTL9, Noe2015KineticSimulation} In this work, we instead employ state-free reversible VAMPnets (SRVs) that can be conceived of as a nonlinear version of tICA. \citep{Chen} SRVs employ neural networks to learn flexible nonlinear functions of the trajectory featurization that better approximate the slow dynamical modes of the system and have been shown to produce substantially higher resolution MSMs than those developed using tICA in applications to WW-domain and Trp-cage mini-proteins.\citep{Chen, Sidky} Importantly, the SRV modes were learned independently for each system to best approximate the slow collective modes for that particular DNA sequence. SRVs were trained using the SRV package we previously developed (\url{https://github.com/hsidky/srv}) employing the default network architecture of two hidden layers each comprising 100 neurons and tanh activation functions, a learning rate of 0.001, and a batch size of 50,000. We adopted a lag time of $\tau$ = 1.2 ns as appropriately short time scale to resolve the dynamical details of the hybridization/dehybridization dynamics. \citep{Phys2018MarkovValidation} As observed by Husic and Pande, it is conceptually incorrect to treat the lag time as a model hyperparameter to be self-consistently optimized via the VAMP-2 score, but rather as a physically-motivated choice designed to expose the dynamical motions relevant at a particular time and length scale of interest. \citep{Husic2017Note:Selection} As we shall show, this choice of lag time leads to converged and Markovian macrostate MSMs. We guarded against overfitting using five-fold cross-validation in which we divided the trajectory data into 50:50 training:validation splits. We observed plateau of the validation loss and no evidence of overfitting after 20 epochs of training for each system that required approximately 22 GPU-minutes on a single NVIDIA Tesla K80 GPU card. A VAMP-2 scoring of the cumulative kinetic variance explained as a function of number of SRV collective modes retained also showed no evidence of overfitting -- as would be evinced by separation of the training and validation VAMP-2 scores \citep{Sidky} -- and exhibited a knee for each of the four DNA sequences after the fifth, fourth, third, and second slow modes for AT-all, GC-end, GC-core, and GC-mix, respectively (\blauw{Fig.~S1} in the \blauw{Supporting Information}), and motivating the construction of 5D, 4D, 3D, and 2D embeddings, respectively.

@@

This material is based upon work supported by the National Science Foundation under Grant No.\ CHE-1841805. This work was completed in part with resources provided by the University of Chicago Research Computing Center. We gratefully acknowledge computing time on the University of Chicago high-performance GPU-based cyberinfrastructure supported by the National Science Foundation under Grant No. DMR-1828629.

@@

%\bibliographystyle{unsrt}
%\bibliographystyle{achemso} % should be pre-defined in usepackage
\bibliography{references_210514}

@@

\begin{figure}[ht!]
	\begin{center} 
        \includegraphics[width=150mm, scale=1]{TOC.pdf}
        %\caption*{TOC image}
        \label{fig:TOC}
	\end{center}
\end{figure}

>>>>

%\usepackage[hidelinks]{hyperref}
%\usepackage{algpseudocode}
%\algdef{SE}[DOWHILE]{Do}{doWhile}{\algorithmicdo}[1]{\algorithmicwhile\ #1}

@@

%\newcommand*\mycommand[1]{\texttt{\emph{#1}}}

%\newcommand{\state}[1]{$\mathcal{S}_{#1}$}
%\newcommand{\Expect}[1]{\mathrm{E}\left[#1\right]}
%\newcommand{\norm}[1]{\left\lVert#1\right\rVert}

%\newcommand*{\rood}[1]{{\color{red}{#1}}}
%\newcommand*{\bood}[1]{{\color{blue}{#1}}}
\newcommand*{\blauw}[1]{\textcolor{blue}{#1}}

%\newcommand*{\addref}[1]{\textcolor{red}{\{ADD REF: #1\}}}
%\newcommand*{\noter}[1]{\textcolor{red}{[[#1]]}}		% notes 
%\newcommand*{\noteb}[1]{\textcolor{blue}{[[#1]]}}		% notes 
%\newcommand*{\noter}[1]{}					% notes off

@@

Markov state models (MSMs) are a powerful approach to infer long-time kinetic models from short molecular simulation trajectories \citep{Sengupta2019AutomatedSelf-assembly, Pande2010EverythingAsk, Chodera2014MarkovDynamics, Husic2018MarkovScience, Wehmeyer2019IntroductionSoftware} that we employ to construct high-resolution sequence-dependent kinetic models of DNA hybridization and dissociation. In brief, a MSM extracts from simulation trajectories an ensemble of long-lived metastable macrostates, their equilibrium occupancy populations, and the equilibrium probability fluxes between them. In this manner, they provide an interpretable and predictive model of the system thermodynamics and long-time kinetics. MSMs have recently been implemented to study the hybridization mechanism of one particular 14-mer DNA oligomer, but determining the sequence-dependent kinetics and mechanisms was not the focus of this study.\citep{Jin2019} An energy disconnectivity graph-based approach was used to interrogate the differences in hybridization rates and mechanisms between GGGGGG and GCGCGC hexamers to reveal strong deviations from ``all-or-nothing'' behaviors and the importance of zippering and slithering mechanisms.~\citep{Xiao2019} Kinetic models were constructed not from the dynamical trajectories of the molecular model, but by estimating rate constants between local minima were estimated using a transition state theory approximation. A recent application of MSMs to the long-time dynamics of short RNA oligonucleotides revealed stacking time scales to be highly sequence dependent.~\citep{Pinamonti2017} In this work, MSMs were constructed for each of the four DNA sequences from the 40$\times$25 $\mu$s simulation trajectories following the six-step protocol detailed in Ref.~\citep{Sidky2019High-ResolutionVAMPnets}: (i) trajectory featurization, (ii) dimensionality reduction, (iii) microstate clustering and microstate transition matrix inference, (iv) macrostate clustering and macrostate transition matrix inference. Calculations were performed using the PyEMMA software package. \citep{Scherer2015PyEMMAModels}

@@

\textbf{Dimensionality reduction.} The featurized trajectories were then projected into a low-dimensional space in preparation for microstate clustering. The standard approach to doing so is to employ time-lagged independent components analysis (tICA) to learn a linear projection into a low-dimensional embedding that maximally preserves the kinetic variance in the data.~\citep{Perez-Hernandez2013IdentificationConstruction, Schwantes2013ImprovementsNTL9, Noe2015KineticSimulation} In this work, we instead employ state-free reversible VAMPnets (SRVs) that can be conceived of as a nonlinear version of tICA. \citep{Chen} SRVs employ neural networks to learn flexible nonlinear functions of the trajectory featurization that better approximate the slow dynamical modes of the system and have been shown to produce substantially higher resolution MSMs than those developed using tICA in applications to WW-domain and Trp-cage mini-proteins.\citep{Chen, Sidky2019High-ResolutionVAMPnets} Importantly, the SRV modes were learned independently for each system to best approximate the slow collective modes for that particular DNA sequence. SRVs were trained using the SRV package we previously developed (\url{https://github.com/hsidky/srv}) employing the default network architecture of two hidden layers each comprising 100 neurons and tanh activation functions, a learning rate of 0.001, and a batch size of 50,000. We adopted a lag time of $\tau$ = 1.2 ns as appropriately short time scale to resolve the dynamical details of the hybridization/dehybridization dynamics. \citep{Phys2011MarkovValidation} As observed by Husic and Pande, it is conceptually incorrect to treat the lag time as a model hyperparameter to be self-consistently optimized via the VAMP-2 score, but rather as a physically-motivated choice designed to expose the dynamical motions relevant at a particular time and length scale of interest. \citep{Husic2017Note:Selection} As we shall show, this choice of lag time leads to converged and Markovian macrostate MSMs. We guarded against overfitting using five-fold cross-validation in which we divided the trajectory data into 50:50 training:validation splits. We observed plateau of the validation loss and no evidence of overfitting after 20 epochs of training for each system that required approximately 22 GPU-minutes on a single NVIDIA Tesla K80 GPU card. A VAMP-2 scoring of the cumulative kinetic variance explained as a function of number of SRV collective modes retained also showed no evidence of overfitting -- as would be evinced by separation of the training and validation VAMP-2 scores \citep{Sidky2019High-ResolutionVAMPnets} -- and exhibited a knee for each of the four DNA sequences after the fifth, fourth, third, and second slow modes for AT-all, GC-end, GC-core, and GC-mix, respectively (\blauw{Fig.~S1} in the \blauw{Supporting Information}), and motivating the construction of 5D, 4D, 3D, and 2D embeddings, respectively.

@@

This material is based upon work supported by the National Science Foundation under Grant No.\ CHE-1841805. B.A. acknowledges support from the NSF Graduate Research Fellowship Program. This work was completed in part with resources provided by the University of Chicago Research Computing Center. We gratefully acknowledge computing time on the University of Chicago high-performance GPU-based cyberinfrastructure supported by the National Science Foundation under Grant No. DMR-1828629.

@@

%\bibliographystyle{unsrt}
%\bibliographystyle{achemso} % should be pre-defined in usepackage
\bibliography{references_210514_abv}

@@

\begin{figure}[ht!]
	\begin{center} 
        \includegraphics[width=150mm, scale=1]{TOC.pdf}
        \caption*{For Table of Contents Only}
        \label{fig:TOC}
	\end{center}
\end{figure}
